Kleene Categories

### 2025-07-10 kleenecat/kleenecat.tex

\documentclass[acmsmall,anonymous,review,screen]{acmart}
\usepackage{mathpartir}
\usepackage{quiver}
\usepackage{enumitem}
\usepackage{wrapfig}
\usepackage{fancyvrb}
\usepackage{amsmath}

@@

\newcommand{\Set}{\cat{Set}}
\newcommand{\lto}{\multimap}
\newcommand{\grammar}{\Set^{\cat{\Sigma^*}}}
\newtheorem*{construction}{Construction}

@@

Let $\cat{C}$ be a category. It is more or less straightforward to
define the free Kleene category $PKC(\cat{C})$. We can do that by
using the internal language of Kleene categories. Concretely, it
will be a type theory with (non-commutative) tensors, sum types and
a primitive type of lists $L(A)$ for every $A$. By type theoretic reasons, this
structure forces the coproduct to distribute over the tensor, where we
can write the morphism $A \otimes (B + C) \to (A \otimes B) + (A
\otimes C) $ as the term
\begin{align*}
  &z \vdash \mathsf{let}\, x \otimes y = z\, \mathsf{in}\\
  &\mathsf{case}\, y \, \mathsf{with}\\
  &| \mathsf{in}_1\, y_1 \Rightarrow \mathsf{in}_1\, (x \otimes y_1)\\
  &| \mathsf{in}_2\,y_2\Rightarrow \mathsf{in}_2 \,(x \otimes y_2)\\
  &\mathsf{end}
\end{align*}
%
The list type includes primitive terms for $\textrm{cons}$,
$\textrm{nil}$, $\textrm{foldr}$ and $\textrm{foldl}$ where
$\textrm{foldr}$ and $\textrm{foldl}$ allow for a context to be used
in the construction of the nil case, but no free variables can be used
in the cons case:
\begin{mathpar}
  \inferrule
  {\Gamma \vdash L : A^* \and \Delta \vdash N : B \and a:A,b:B \vdash M : B
  }
  {\Gamma,\Delta \vdash \textrm{foldr}(L,N,a.b.M) : B}
\end{mathpar}

@@

\section{Gluing Kleene categories}

The following theorems should be true
\begin{theorem}
  Given Kleene categories $\cat{L}$, $\cat{K}$, $\cat{K}'$
  and functors $F : \cat{L} \to \cat{K}$, $p : \cat{K}' \to \cat{K}$
  such that $F$ is lax monoidal such that it laxly preserves the initial
  algebras and $p$ is a monoidal fibration strictly preserving the
  Kleene category structure, then in the pullback diagram
  % https://q.uiver.app/#q=WzAsNCxbMCwxLCJcXGNhdHtMfSJdLFsxLDEsIlxcY2F0e0t9Il0sWzEsMCwiXFxjYXR7S30nIl0sWzAsMCwiXFxjYXR7TH0nIl0sWzAsMSwiRiIsMl0sWzIsMSwicCJdLFszLDJdLFszLDBdLFszLDEsIiIsMSx7InN0eWxlIjp7Im5hbWUiOiJjb3JuZXIifX1dXQ==
\[\begin{tikzcd}
	{\cat{L}'} & {\cat{K}'} \\
	{\cat{L}} & {\cat{K}}
	\arrow[from=1-1, to=1-2]
	\arrow[from=1-1, to=2-1]
	\arrow["\lrcorner"{anchor=center, pos=0.125}, draw=none, from=1-1, to=2-2]
	\arrow["p", from=1-2, to=2-2]
	\arrow["F"', from=2-1, to=2-2]
\end{tikzcd}\]
$\cat{L}'$ is a Kleene category and the functor $\cat{L}'\to \cat{L}$ is a
monoidal fibration strictly preserving the Kleene category structure.
\end{theorem}
\begin{proof}
  It is known from linear logic reasons that the pullback is distributive monoidal.
  The Kleene star should be defined much like the exponential in glued models of
  linear logic. And the projection functor is a monoidal fibration because those
  are stable under pullback.
\end{proof}

Recently I have been enjoying comma categories a bit more, which leads me to the follwing
conjecture:

\begin{conjecture}
  The 2-category of Kleene categories has comma objects.
\end{conjecture}

>>>>

\documentclass[acmsmall,anonymous,review,screen]{acmart}
\usepackage{mathpartir}
\usepackage{quiver}
\usepackage{enumitem}
\usepackage{wrapfig}
\usepackage{fancyvrb}
\usepackage{amsmath}
\usepackage{stmaryrd}

@@

\newcommand{\FKC}{\textrm{F}_{\textrm{KC}}}
\newcommand{\sem}[1]{\llbracket{#1}\rrbracket}
\newcommand{\Set}{\cat{Set}}
\newcommand{\lto}{\multimap}
\newcommand{\grammar}{\Set^{\cat{\Sigma^*}}}
\newcommand{\id}{\textrm{id}}
\newtheorem*{construction}{Construction}

@@

Let $\cat{C}$ be a category. It is more or less straightforward to
define the free Kleene category $\FKC(\cat{C})$. We can do that by
using the internal language of Kleene categories. Concretely, it
will be a type theory with (non-commutative) tensors, sum types and
a primitive type of lists $L(A)$ for every $A$. By type theoretic reasons, this
structure forces the coproduct to distribute over the tensor, where we
can write the morphism $A \otimes (B + C) \to (A \otimes B) + (A
\otimes C) $ as the term
\begin{align*}
  &z \vdash \mathsf{let}\, x \otimes y = z\, \mathsf{in}\\
  &\mathsf{case}\, y \, \mathsf{with}\\
  &| \mathsf{in}_1\, y_1 \Rightarrow \mathsf{in}_1\, (x \otimes y_1)\\
  &| \mathsf{in}_2\,y_2\Rightarrow \mathsf{in}_2 \,(x \otimes y_2)\\
  &\mathsf{end}
\end{align*}
%
The list type includes primitive terms for $\textrm{cons}$,
$\textrm{nil}$, $\textrm{foldr}$ and $\textrm{foldl}$ where
$\textrm{foldr}$ and $\textrm{foldl}$ allow for a context to be used
in the construction of the nil case, but no free variables can be used
in the cons case:
\begin{mathpar}
  \inferrule
  {\Gamma \vdash L : A^* \and \Delta \vdash N : B \and a:A,b:B \vdash M : B
  }
  {\Gamma,\Delta \vdash \textrm{foldr}(L,N,a.b.M) : B}
\end{mathpar}

@@

This should imply that the algebraic and coalgebraic semantics of regular grammars coincides?

\section{Canonicity for Free Kleene Categories}

Let $\Sigma$ be a set. Then the ``standard semantics'' of the free
Kleene category $\FKC(\Sigma)$ is the unique functor $\sem{\cdot} :
\FKC(\Sigma) \to \Set^{\Sigma^*}$ preserving the (Day convolution)
Kleene category structure.

We first address whether we can generalize Kozen's completeness
theorems for Kleene algebras. Firstly, the fact that the Free Kleene
category is complete for \emph{language} inclusion and \emph{weak}
equivalence follows immediately from the fact that the free Kleene
algebra is equivalent to the posetal reflection of the free Kleene
category and the completeness theorem for Kleene algebras. That is,
for any objects $A, B \in \FKC(\Sigma)$, if $|\sem{A}| \subseteq
|\sem{B}|$ then there is a morphism $A \to B$ in $\FKC(\Sigma)$.

The immediate generalization to strong equivalence is false. For
example, we might ask that $\sem{\cdot}$ is a \emph{full} functor,
that any semantic reduction of the formal grammars is the denotation
of a morphism in the free Kleene category. But consider that the
regular grammar $1^*$ denotes a discrete grammar of natural numbers
and so morphisms $1^* \otimes \cdots \to 1^*$ are mapped into a set of
multi-argument functions on natural numbers. So completeness would
imply that all functions on natural numbers can be defined in the
syntax of the free Kleene category, which fails as all such functions
should be primitive recursive\footnote{flesh out the
details?}.

Failing completeness for definability of reductions, we could ask that
the \emph{equational theory} of free Kleene categories is
complete. This means that $\sem{\cdot}$ is a \emph{faithful} functor,
i.e., if two syntactic morphisms denote the same reduction $\sem{M} =
\sem{N}$, then they are provably equal $M = N$. I don't know if this
is true yet, but it seems unlikely to me.

However there is a (very) restricted form of completeness that is
true. The denotation $\sem{\cdot}$ is fully faithful when restricting
to hom sets of the form $\FKC(\Sigma)(\lceil w \rceil, A)$. The
significance of this is that the denotation of such a morphism is
equivalent to $\sem{A}\,w$, the set of ``parse trees'' for the grammar
$A$. Then fullness says that the term syntax is at least complete
enough to describe all parse trees, and faithfulness says that the
equational theory is at least complete enough to prove that terms that
represent the same parse tree are provably equal. In other words, for
every parse tree, there is a unique equivalence class of terms that
map to it. This theorem is called canonicity because it says that all
terms of a certain type are provably equal to a term in a
``canonical'' form.

\begin{theorem}{Canonicity}
  The denotation $\sem{\cdot}$ is bijective when restricting to hom
  sets of the form $\FKC(\Sigma)(\lceil w \rceil, A)$.
\end{theorem}

To prove this theorem, we first change perspective. The function
$\lceil \cdot \rceil : \Sigma^* \to \FKC(\Sigma)$ induces a restricted
Yoneda embedding called the ``nerve'' functor $N : \FKC(\Sigma) \to
\Set^{\Sigma^*}$ defined by
\[ N(A)w = \FKC(\Sigma)(\lceil w \rceil, A) \]
The action of $\sem{\cdot}$ on the restricted hom sets then induces a
natural transformation
\[ u : N \Rightarrow \sem{\cdot} \]
defined as
\[ u_w(f : \FKC(\Sigma)(\lceil w \rceil , A)) = \sem{f}(\iota_w) : A w \]
where $\iota_w : \sem{\lceil w \rceil} w$ is the unique parse tree
witnessing that the grammar $Y c_1 \otimes \cdots \otimes \varepsilon$
matches the string $c_1\cdots$. This $\iota_w$ is a universal element
of $\sem{\lceil w \rceil}$, meaning that evaluating at $\iota_w$ is a
bijection between $\Set^{\Sigma^*}(\sem{\lceil w \rceil}, A)$ and $A w$. So we
have that $\sem{\cdot}$ is fully faithful on these hom sets if and
only if $u$ is a natural isomorphism.

To do this we will explicitly construct an inverse natural
transformation $q : \sem{\cdot} \to N$ which we call ``quote'' because it quotes
a semantic parse tree back into the syntax. $q$ can be defined
explicitly by recursion on syntactic types and naturality proven by
induction over syntactic terms but we will develop a more structured
proof using the universal property of the free Kleene category.  To do
this we need to develop a Kleene category for which constructing a
functor from $\FKC(\Sigma)$ into it is equivalent to defining such a
natural transformation. We can do this by using the following
properties of \emph{comma categories}.

\begin{lemma}
  Let $F : \cat C \to \cat E$ and $G : \cat D \to \cat E$ and let $F
  \downarrow G$ be the comma category (i.e., whose objects are
  morphisms $F c \to G d$) with projections $p_C,p_D$.
  \begin{enumerate}
  \item If $\cat C, \cat D, \cat E$ have a monoidal structure for
    which $F$ is oplax and $G$ is lax monoidal, then $F \downarrow G$
    can be equipped with a monoidal structure such that $p_C,p_D$ are
    strict monoidal.
  \item If $\cat C, \cat D$ have (specified) colimits of shape
    $\mathcal I$ and $F$ preserves colimits of shape $\mathcal I$ then
    $F \downarrow G$ has colimits of shape $\mathcal I$ and the
    projections (strictly) preserve them.
  \end{enumerate}
\end{lemma}

Then we can prove the following
\begin{theorem}{Gluing for Kleene categories}
  Let $\cat K, \cat L, \cat M$ be Kleene categories and $\Phi : \cat K
  \to \cat M$ be a Kleene functor and $G : \cat L \to \cat M$ be a lax
  monoidal functor. Then $\Phi \downarrow G$ is a Kleene category, and
  the projections are (strict) Kleene functors.
\end{theorem}
\begin{proof}
  The monoidal and coproduct aspects of Kleene categories are handled
  by the previous lemma. We need to show that the monoidal structure
  is distributive and construct the Kleene star.

  For the Kleene star object, given $f :
  \Phi(k) \Rightarrow G l$ we define the carrier as
  \[ \Phi(k^*) \cong \Phi(k)^* \to (G l)^* \to G l^* \]
  where $f^* : \Phi(k)^* \to (G l)^*$ and $g : (G l)^* \to G l^*$ is
  defined by the induced cons-structure $G l \otimes G l^* \to G (l
  \otimes l^*) \to G l^*$ and $I \to G I \to G l^*$. Then it is
  routine to check that the nil, cons and fold all lift to commutative
  squares.

  TODO: address the snoc-part as well, and distributivity
\end{proof}

Then $\sem {\cdot} : \FKC(\Sigma) \to \Set^{\Sigma^*}$ is a strict
Kleene functor and $N : \FKC(\Sigma) \to \Set^{\Sigma^*}$ is a lax
monoidal functor, and so $\sem{\cdot}\downarrow N$ is a Kleene
category and the projection to $\FKC(\Sigma)^2$ is a strict Kleene
functor. Then by the universal property of $\FKC(\Sigma)$, there is a
unique strict Kleene functor $q$ from $\FKC(\Sigma)$ to
$\sem{\cdot}\downarrow N$ and further the composition with the
projection to $\FKC(\Sigma)^2$ is strict Kleene and so $q$ is
equivalent to a natural transformation $\sem{\cdot} \Rightarrow N$.

\begin{lemma}
  $q \circ u$ is the identity
\end{lemma}
\begin{proof}
  Let $f : \FKC(\Sigma)(\lceil w \rceil, A)$, then by naturality,
  $q(\sem{f}(\iota)) = f \circ q(\sem{\id}(\iota))$ so it suffices to
  show that $q(\iota) = \id$. This can be confirmed by calculation by
  induction on the length of $w$.
\end{proof}

\begin{lemma}
  $u \circ q$ is the identity
\end{lemma}
\begin{proof}
  By the universal property of $\FKC(\Sigma)$, it suffices to show
  that $u \circ q$ and $\id$ correspond to (strict?) Kleene functors
  from $\FKC(\Sigma)$ to $\sem{\cdot} \downarrow \sem{\cdot}$. The
  identity part should be a general theorem, and verifying the Kleene
  category properties is done by cases.
\end{proof}
